---
title: "SST encounters tidyverse"
subtitle: "Some doodle for PGNAPES"
author: "Ziggy and the Spider"
output: 
  html_document: 
    fig_height: 7.0
    fig_width: 9.5
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE)
```

```{r}
library(raster)
library(tidyverse)
# devtools::install_github("einarhjorleifsson/gisland", dependencies = FALSE)
```

TODO: Include documentation of downloading from source. For now check the [nco-scripts directory](https://github.com/sigurdurthorjonsson/myoisst/tree/master/nco-scripts) on Ziggy's repo. 

TODO2: Similar plots for May and June have been requested by members of the survey group. A move of ncdf-files to the Hafró-ftp-server would then be natural.

Read in data:
```{r import}
years <- 2010:2018

# seeding
res <- list()

# stack the hay
for(i in seq(along = years)) {
  res[[i]] <- 
    raster(paste("macArea", years[i], 
                 "m07.nc", sep = "/"), varname = "sst") %>% 
    # User can change spatial resolution of temperature grid in the next step.
    #  Higher values mean faster plot rendering.
    aggregate(fac = 2) %>% 
    as("SpatialPixelsDataFrame") %>% 
    as.data.frame() %>% 
    as_tibble() %>% 
    rename(sst = Daily.sea.surface.temperature) %>% 
    mutate(year = years[i])
}

# bind the hay
d <-
  res %>% 
  bind_rows()
```

What have we got?:
```{r}
glimpse(d)
```


Get a background map of terrestrial reference areas:
```{r map}
m <- map_data("world", xlim = c(-50, 50), ylim = c(30, 85))
```

The sst plot (Be patient, takes some time to render. Or worse, if you got a lousy computer you will have a meltdown. In that case ask your boss for a real a computer that can do some modern day number crunching):
```{r sstplot}
# Find plot range
xlim <- range(d$x)
ylim <- range(d$y)

# The number of "steps" in the rainbow:
nlev <- d$sst %>% floor() %>% unique() %>% length()

d %>% 
  ggplot() +
  theme_bw() +
  geom_tile(aes(x=x, y=y, fill=sst), alpha=0.8) +
  scale_fill_gradientn(colours = rainbow(nlev)) +
  geom_polygon(data = m, aes(long, lat, group = group), fill = "grey") +
  coord_map(projection = "azequalarea",
            xlim = xlim, ylim = ylim) + 
  scale_x_continuous(name = NULL, breaks = NULL) +
  scale_y_continuous(name = NULL, breaks = NULL) +
  labs(title = "July - average SST",
       fill = "°C") +
  facet_wrap(~ year)
```

Calculate annual anomalies by "square":
```{r calcanomalies}
d <- 
  d %>% 
  group_by(x, y) %>% 
  mutate(anomaly = sst - mean(sst)) %>% 
  ungroup()
```


Check the anomaly distribution:
```{r distribution}
d %>% 
  mutate(anomaly = gisland::grade(anomaly, 0.5)) %>% 
  group_by(year, anomaly) %>% 
  count() %>% 
  ggplot(aes(anomaly, n)) +
  geom_col() +
  facet_wrap(~ year)
```

Anomalies beyond c(-3, 3) are all but extremes (at least at the aggregation of 2 on the raw data), hence for the plotting lets trim the extremes:
```{r trimming}
amin <- -3
amax <- 3
d <-
  d %>% 
  # "Trim" the exremes, used in the plotting
  mutate(a.plot = case_when(anomaly < amin ~ amin,
                       anomaly > amax ~ amax,
                       TRUE ~ anomaly),
         a.plot = gisland::grade(a.plot, 0.5))
```

And the anomaly plot: NOTE this referenced to 2010--2018. (Again, be patient, takes some time to render):
```{r anomalyplot}
d %>%
  ggplot() +
  # lets keep the default theme, shows the extent of the data
  #theme_bw() +
  geom_tile(aes(x=x, y=y, fill= round(anomaly)), alpha=0.8) +
  scale_fill_gradient2(low = "blue", mid = "white", high = "red",
                       midpoint = 0, limits = c(-3, 3),
                       breaks = seq(-3, 3, by = 0.5),
                       guide = "legend") +
  geom_polygon(data = m, aes(long, lat, group = group), fill = "grey") +
  coord_map(projection = "azequalarea", xlim = xlim, ylim = ylim) + 
  scale_x_continuous(name = NULL, breaks = NULL) +
  scale_y_continuous(name = NULL, breaks = NULL) +
  labs(title = "July SST anomaly",
       fill = "Anomaly +/- 0.25") +
  facet_wrap(~ year)
```

And just a little sanity check on values vs colours:
```{r sanitycheck}
amin <- -5
amax <-  5
set.seed(314) # for the love of pi
d <- 
  expand.grid(x = 1:10,
              y = 1:10) %>% 
  as_tibble() %>% 
  mutate(z = runif(n(), -1, 10),
         anomaly = z - mean(z),
         a.plot = case_when(anomaly < amin ~ amin,
                            anomaly > amax ~ amax,
                            TRUE ~ anomaly),
         a.plot = gisland::grade(a.plot, 0.5))
d %>% 
  ggplot() +
  theme_bw() +
  geom_tile(aes(x = x, y = y, fill = a.plot)) +
  geom_text(aes(x = x, y = y, label = round(anomaly, 2))) +
  scale_fill_gradient2(low = "blue", mid = "white", high = "red",
                       midpoint = 0, limits = c(-5, 5),
                       breaks = seq(-5, 5, by = 0.5),
                       guide = "legend")
```

